---
title: "Snowshoe Hares"
author: "Ian Brunjes"
date: "11/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(tidyverse)
library(ggbeeswarm)
library(kableExtra)
library(here)
library(effsize)
library(broom)
```

## 1. Introduction

Snowshoe hares are a keystone prey species in the northern boreal forests of Alaska^1^ and therefore represent a vital indicator of overall habitat and population health. A scientific initiative pursued at Bonanza Creek Experimental Forest has yielded insights into several attributes of these hare populations.

This document will provide an exploratory analysis of these observations on snowshoe hares taken in the Tanana Valley from 1999-2012. Specifically, we examine the differences in weight between male and female juvenile snowshoe hares, the relationship between weight and hind foot length, and overall trends in observation counts across the span of the study.

## 2. Data & Analyses

This data is the result of capture-recapture studies of snowshoe hares at 5 locales in the Tanana valley, from Tok in the east to Clear in the west in the years from 1999 to 2002. Physical attributes including sex, age, weight, and hind foot size were recorded against the trapped hares, as well as site and trap identifiers of the observations.

Using summary statistics, 2-sample t-tests, and linear regression modeling, we compare the juvenile weights of male and female snowshoe hares, as well as the relationship between weight and hind foot length.

Data is made available by the Bonanza Creek LTER and the Environmental Data Initiative.^1^

```{r include=FALSE}
# Read in data from 'data/bonanza_hares.csv'
hares_data <- read_csv(here("data", "bonanza_hares.csv"))
```

## 3. Exploratory Findings

### 3.1 Juvenile hare trap counts

```{r include=FALSE}
# Filter hare data to examine only juvenile hare traps
juvenile_hares <- hares_data %>%
  filter(age == "j")

# Coerce the date to a more parse-able format to derive year
juvenile_hares_year <- juvenile_hares %>% 
  mutate(date = mdy(date)) %>% 
  mutate(year = year(date))
  
# Get the total counts of hares recorded in each year of the study
juvenile_hares_annual_count <- juvenile_hares_year %>% 
  group_by(year) %>% 
  count(name = "count")
```

```{r echo=FALSE}
# Plot the annual counts of juvenile hares
ggplot(data = juvenile_hares_annual_count, aes(x = year, y = count)) +
  geom_col(fill="#f68060") +
  labs(y = "Number of Hares Observed\n") +
  scale_x_continuous(breaks = seq(1999, 2012, 1)) +
  scale_y_continuous(breaks = seq(0, 120, 20)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    axis.title.x = element_blank(),
    axis.title = element_text(size=12),
    axis.text = element_text(size=10, face="bold")
  )
```

**Figure 3.1.1:** Annual Hare Counts.

```{r include=FALSE}
# Find max/mean/median values of annual hare counts
annual_count_mean <- mean(juvenile_hares_annual_count$count)
annual_count_median <- median(juvenile_hares_annual_count$count)
annual_count_max <- max(juvenile_hares_annual_count$count)
```

There is a general downward trend in the number of hares counted over time. The maximum number of hares (`r annual_count_max`), observed in the very first year, was much higher than the mean (`r annual_count_mean`) or median (`r annual_count_median`) observed across all years. To standardize the hare population estimates based on this count data, I would suggest creating a metric of hares counted per traps set per day. Since the number of traps utilized in each site and number of days visited in a site could skew the number of hares observed to be higher than in other sites.

### 3.2 Juvenile hare weights

```{r echo=FALSE, warning=FALSE}
# Build label vectors for plot annotation
site_names <- c("Black Spruce", "Mature", "Riparian")
names(site_names) <- c("bonbs", "bonmat", "bonrip")
sexes <- c("Female", "Male", "NA")

# Plot hare weight and sex for each site
ggplot(data = juvenile_hares, aes(x = sex, y = weight)) +
  geom_boxplot(width = 0.5, outlier.color = NA) +
  geom_point(aes(color = sex), alpha = 0.3, size = 2) +
  facet_wrap(~grid, labeller = labeller(grid = site_names)) +
  theme_light() +
  labs(x = "\nSex",
       y = "Weight (g)\n") +
  theme(legend.position = "none") +
  scale_x_discrete(labels = sexes)
```

**Figure 3.2.1:** Weight of juvenile hares by sex at each observation site.

The observations of female and male snowshoe hare weight, depicted within each site where they were recorded, shows there is a relative similarity between the weights of the two sexes. From the box plot summary statistics, we can notice that while the median weight for males was higher at each site, the interquartile range of male and female at each site overlap quite a bit.


### 3.3 Juvenile weight comparison

**Table 3.3.1:** Descriptive statistics (mean, median, standard deviation, and sample size) of the weight of juvenile male and female hares.
```{r echo = FALSE, message=FALSE}
# Build summary statistics table for weight of male and female juvenile hares
juvenile_hares_stats <- juvenile_hares %>%
  drop_na(sex) %>% # only comparing male and female sex specifically, NA not needed
  drop_na(weight) %>%  # NA values won't contribute to summary stats, so remove for accurate sample size
  mutate(sex = case_when(
    sex == "f" ~ "Female",
    sex == "m" ~ "Male"
  )) %>% 
  group_by(sex) %>% 
  summarize(
    mean = mean(weight),
    median = median(weight),
    sd = sd(weight),
    count = n()
  )

# Output summary stats as nicely formatted table
juvenile_hares_stats %>% 
  kable(col.names = c("Sex", "Mean weight (g)", "Median weight (g)", "Standard deviation (g)", "Sample size")) %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r include=FALSE}
# Derive our two samples for weight comparison: male and female hares
female_juvenile_weights <- juvenile_hares %>% 
  filter(sex == "f") %>%
  drop_na(weight) %>% 
  pull(weight)

male_juvenile_weights <- juvenile_hares %>% 
  filter(sex == "m") %>% 
  drop_na(weight) %>% 
  pull(weight)

# Plot histograms & qq-plots to check assumptions
hist(male_juvenile_weights) # Looks relatively normal
qqnorm(male_juvenile_weights) # Relatively linear
hist(female_juvenile_weights) # Looks normal
qqnorm(female_juvenile_weights) # Looks very linear

# Overall: assumptions of normality hold, & sample size additionally >> Central Limit Theorem (means will be normally distributed). Two-sample t-test OK for means comparison!

# Run 2-sample t-test using these vectors
t.test(x = female_juvenile_weights, y = male_juvenile_weights)

# Get the means & SDs for each: 
male_mean <- mean(male_juvenile_weights)
female_mean <- mean(female_juvenile_weights)
male_sd <- sd(male_juvenile_weights)
female_sd <- sd(female_juvenile_weights)

# Calculate percent difference of means
mean_diff_pct <- round((male_mean-female_mean)/female_mean * 100, 2)

# Get the effect size: 
hares_effsize <- cohen.d(female_juvenile_weights, male_juvenile_weights)
```

From our t-test, we see the mean weight for the male hares is `r mean_diff_pct`% higher than mean size for the female group.

If these samples are indeed drawn from populations that have the same mean size (which is our null hypothesis of this t-test), there is a 0.7% chance that two random samples from these populations would result in samples that are at least as different as the ones observed here. What this means is that if there was no difference in weight between male and female samples, there is only a 0.7% chance that a random sampling of this population would result in a difference in mean that is at least as much as we have here.

While the p-value might indicate significance, our effect size is on the small side (Cohen's *d* = `r round(hares_effsize$estimate,2)`).

### 3.4 Juvenile weight vs hind foot length

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Massage the data set for comparison of weight vs hind foot length
juvenile_hares %>% ggplot(aes(x = weight, y = hindft)) +
  geom_point(color = "cadetblue") +
  theme_minimal() +
  geom_smooth(method = "lm", se = FALSE, color = "gray30") +
  labs(x = "\nWeight (g)", y = "Hind foot length (mm)\n") +
  theme(legend.position = c(0.8,0.25))
```

**Figure 3.4.1:** Relationship of weight (g) and hind foot length (mm) in juvenile snowshoe hares.

```{r include=FALSE}
# Build a linear regression model for weight and height foot length relation
hares_lm <- lm(hindft ~ weight, data = juvenile_hares)
hares_lm_tidy <- broom::tidy(hares_lm)

hares_int <- hares_lm_tidy$estimate[[1]]
hares_coef <- hares_lm_tidy$estimate[[2]]

hares_lm_out <- broom::glance(hares_lm)

# Check assumptions for linear regression usage using plot()
plot(hares_lm)
# The spread of residuals about the fitted value looks fairly constant across the span of the x axis.
# Evidence of homostedasticity
# No outliers identified with too much influence in the model
# The residuals appear normally distributed except at the beginning which might indicate an issue in attempting to apply a linear model

# Get Pearson's r correlation
hares_cor <- cor.test(juvenile_hares$hindft, juvenile_hares$weight)

# Tidy version of correlation output:
hares_cor_tidy <- tidy(hares_cor)
```

Simple linear regression was used to explore the relationship between penguin weight (g) and hind foot length (mm) for all recorded juvenile snoeshow hares. A significant regression model was found ($\beta$ = `r round(hares_coef,3)`, F(`r hares_lm_out$df`,`r hares_lm_out$df.residual`) = `r round(hares_lm_out$statistic,1)`, p < 0.001) with an R^2^ of `r round(hares_lm_out$r.squared,3)`.

The relationship between weight and hind foot length appears relatively linear (Figure 3.4.1). Simple linear regression revealed that weight only somewhat predicts hind foot length (p < 0.001, R^2^ = `r round(hares_lm_out$r.squared,2)`) with an average slope of $\beta$ = `r round(hares_lm_tidy$estimate[2], 2)` mm g^-1^ (i.e., for each one gram increase in weight we expect an average increase in weight of `r round(hares_lm_tidy$estimate[2], 2)` mm). Hare weight and hind foot length have a moderate positive correlation (Pearson's *r* = `r round(hares_cor_tidy$estimate,2)`, p < 0.001).


Checking against the assumptions of linear regression, there is evidence of homoscedastic residuals with no strongly influencing outliers, but there is a concerning trend in lower weight values that might indicate the linear model is not accurately descriptive of the relationship between weight and hind foot length in juvenile snowshoe hares.


## 4. Summary

Exploratory data analysis reveals the following initial findings: 

- There is a decrease in snowshoe hare population being observed from 1999 - 2012.
- In juvenile snowshoe hares, males are observed to have higher weights than females (the difference is significant, and the effect size is small).
- In juvenile snowshoe hares, Weight appears linearly related with hind foot length; the relationship (by simple linear regression) is significant, with a slope (expected average increase in hind foot length with a 1 g increase in weight) of `r round(hares_lm_tidy$estimate[2], 2)` mm g^-1^.

## 5. Citations

^1^**Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER.** 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14

